---
title: "Working with SNP Data"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Quick Filtering and Management of Array Data
This actually happens outside of R.  A cool feature of RStudio

### Filter by missing genotypes, missing individuals and create a VCF File

```{bash, eval=FALSE}
plink2 --bfile ../Example_Data/final --recode vcf id-paste=iid --out ../Processed_Data/array --allow-extra-chr --mind 0.1 --geno 0.05 --max-alleles 2 --min-alleles 2 --chr 1-10
```

Here, we are using the program `plink` to convert the files we got from the genotyping service (`final.ped`, `final.bed`, `final.fam`).  This command has several options:

`--bfile ../Example_Data/final` this is saying the input is a binary and `../Example_Data/final` is a relative PATH to the data files.  In the virtual machine, this could aslo be written as `/home/eogwparticipant/NACE_MAS_Genomics_Workshop/Example_Data` or `~/NACE_MAS_Genomics_Workshop/Example_Data`

`--recode vcf id-paste=iid` is telling it to recode the file as a VCF file using the "individual id" from the `.fam` file as the individual label

`--out array` is telling it to name the output files with the prefix "array"

`--allow-extra-chr` is a Plink specific setting needed when you're not working with human data

#### Important part
`--mind 0.1` This is saying we only want individuals with less than 10% missing data
`--geno 0.05` This is saying we only want SNPs with less than 5% missing data
`--max-alleles 2 --min-alleles 2` This says we only want SNPs with two alleles
`--chr 1-10` This says we only want nuclear SNPs.  The array has other markser on it including mtDNA and pathogens, "chr 1-10" means only the SNPs on the 10 oyster genome chromosomes


It will help us to have both a `.bed` file and `.vcf` later one.  The code below does the same as the above, but outputs a `.bed` file instead
```{bash, eval=FALSE}
plink2 --bfile ../Example_Data/final --make-bed  --mind 0.1 --geno 0.05 --max-alleles 2 --min-alleles 2 --chr 1-10 --allow-extra-chr --out ../Processed_Data/array
```




## Import data into R

Now we're ready to import that data into R for processing.  We will first use the library [vcfR](https://cran.r-project.org/web/packages/vcfR/index.html) to do this
```{r}
library(vcfR)
my_vcf <- read.vcfR("../Processed_Data/array.vcf")
```

Above, we load the library and then use a function to read our vcf and store it as `my_vcf`

## Calculating heterozygosity

`.vcf` and `.bed` files do not contain information about populations or localities.  We need to tell our analyses which individuals below to which populations.  To do this, one was is to use a simple text table.  We can read the one I made for this data like so:

```{r}
table <- read.table("../Example_Data/strata", header = TRUE)
head(table)

poplist.names <- table$Population
```

As you can see our data has 45 individuals total, 15 in PopA, 15 in PopB, and 15 in PopC.


Next, the package vcfR can actually calculate heterozygosity for our data using the `genetic_diff` function

```{r}
het_results <- genetic_diff(my_vcf, pop=as.factor(poplist.names), method= 'nei')

head(het_results)
```

You can see that this dataframe has the value calculated per SNP (each row).  We can take a look at the columns that we are interested in and then take the mean values to get the mean genome-wide values

```{r}
head(het_results[,3:6])
```

```{r}
round(colMeans(het_results[,c(3:6)], na.rm = TRUE), digits = 3)
```

Hs_PopA is the observed heterozygostiy for PopA and so on.  Ht is the observed across all individuals.

### Plotting heterozygosity
I'm a fan of ggplot, but in order to use ggplot we need data in a "tidy" format.  We can use the package reshape to do this for us

```{r}
library(reshape2)
het_df <- melt(het_results[,c(3:6)], varnames=c('Index', 'Sample'), value.name = 'Heterozygosity', na.rm=TRUE)
head(het_df)
```


Now we can easily plot with ggplot.  I'm going to use a simple box plot.

```{r}
p <- ggplot(het_df, aes(x=variable, y=Heterozygosity)) + geom_boxplot(fill="#1F78B4")
p <- p + xlab("Population")
p <- p + ylab("Observed Heterozygosity")
p <- p + theme_bw() + ggtitle("Observed Heterozygosity by population and total")
p
```

## Calculating *F<sub>IS</sub>* (Inbreeding Coeffecient)
```{r}
library(adegenet)
my_genind <- vcfR2genind(my_vcf)
strata<- read.table("../Example_Data/strata", header=TRUE)
strata_df <- data.frame(strata)

strata(my_genind) <- strata_df

setPop(my_genind) <- ~Population
```

```{r}
library(hierfstat)
basicstat <- basic.stats(my_genind, diploid = TRUE, digits = 3)
```

```{r}
summary(basicstat)
head(basicstat$Fis)
fis_df <- melt(basicstat$Fis, varnames=c('Index', 'Sample'), value.name = 'FIS', na.rm=TRUE)
```
```{r}
p <- ggplot(fis_df, aes(x=Sample, y=FIS)) + geom_boxplot(fill="#1F78B4", notch= FALSE)
p <- p + xlab("Population")
p <- p + ylab("FIS")
p <- p + theme_bw() + ggtitle("FIS by population")
p
```

## PCA and DAPC

One important consideration is that we need to deal with linkage in our data set we will use PCAdapt to help us with this and to illustrate why.


```{r}
library(pcadapt)
filename <- read.pcadapt("../Processed_Data/array.bed", type = "bed")

```

```{r}

res <- pcadapt(filename, K = 5)
plot(res, option = "screeplot")
```


```{r}
par(mfrow = c(2, 2))
for (i in 1:4)
  plot(res$loadings[, i], pch = 19, cex = .3, ylab = paste0("Loadings PC", i))
```

```{r, warning=FALSE}
plot(res, option = "scores", pop = poplist.names)
```


```{r}
res <- pcadapt(filename, K = 10, LD.clumping = list(size = 500, thr = 0.2))
plot(res, option = "screeplot")
```

```{r}
res <- pcadapt(filename, K = 4, LD.clumping = list(size = 500, thr = 0.2))
par(mfrow = c(2, 2))
for (i in 1:4)
  plot(res$loadings[, i], pch = 19, cex = .3, ylab = paste0("Loadings PC", i))
```


```{r, warning = FALSE}
plot(res, option = "scores", pop = poplist.names)
```



## Filter data for PCA and DAPC
```{r}
ld_filtered_vcf <- my_vcf[!is.na(res$loadings[,1]),]
mygl <- vcfR2genlight(ld_filtered_vcf,n.cores =2)

strata(mygl) <- strata_df

setPop(mygl) <- ~Population


```

```{r}
oyster.pca <- glPca(mygl, nf = 3)

```


```{r}
barplot(100*oyster.pca$eig/sum(oyster.pca$eig), col = heat.colors(50), main="PCA Variance Explained")
title(ylab="Percent of variance\nexplained", line = 2)
title(xlab="PC", line = 1)

```


```{r}

myCol <- colorplot(oyster.pca$scores,oyster.pca$scores, transp=TRUE, cex=4)
abline(h=0,v=0, col="grey")
add.scatter.eig(oyster.pca$eig[1:10],2,1,2, posi="bottomright", ratio=.16)
```

```{r}

oyster.pca.df <- as.data.frame(oyster.pca$scores)
oyster.pca.df$pop <- pop(mygl)
cols <-c("#871a1a","#33A02C","#1F78B4")

p <- ggplot(oyster.pca.df, aes(x=PC1, y=PC2, colour=pop)) 
p <- p + geom_point(size=2)
p <- p + stat_ellipse(level = 0.95, size = 1)
p <- p + scale_color_manual(values = cols) 
p <- p + geom_hline(yintercept = 0) 
p <- p + geom_vline(xintercept = 0) 
p <- p + theme_bw()

p
```

```{r}
grp <- find.clusters(mygl, max.n.clust=10)
table(pop(mygl), grp$grp)
```

```{r}
oyster.dapc <- dapc(mygl, n.pca = 3, n.da = 2, grp$grp)
scatter(oyster.dapc,col=cols,bg="white", solid=1)
```

```{r}
compoplot(oyster.dapc, posi="bottom",txt.leg=paste("Cluster", 1:6), lab="", ncol=1, xlab="individuals")
```

```{r}
```

```{r}
oyster.dapc <- dapc(mygl, n.pca = 3, n.da = 2)
scatter(oyster.dapc, col = cols, cex = 2, legend = TRUE, clabel = F, posi.leg = "bottomleft", scree.pca = TRUE,
        posi.pca = "topleft", cleg = 0.75, posi.da = "topright")
```

```{r}
compoplot(oyster.dapc, posi="bottom",txt.leg=paste("Cluster", 1:3), lab="", ncol=1, xlab="individuals")
```
```{r}
dapc.df <- as.data.frame(oyster.dapc$posterior)
dapc.df$pop <- pop(mygl)
dapc.df$indNames <- rownames(dapc.df)

library(tidyr)
dapc.df <- pivot_longer(dapc.df, -c(pop, indNames))
colnames(dapc.df) <- c("Original_Pop","Sample","Assigned_Pop","Posterior_membership_probability")

p <- ggplot(dapc.df, aes(x=Sample, y=Posterior_membership_probability, fill=Assigned_Pop))
p <- p + geom_bar(stat='identity') 
p <- p + scale_fill_manual(values = cols) 
p <- p + facet_grid(~Original_Pop, scales = "free")
p <- p + theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 8))
p <- p + ylab("Posterior membership probability")
p
```



```{r}
adegenetServer(what = "DAPC")
```




